
---
title: 'Towards quantitative DNA Metabarcoding: A method to overcome PCR amplification bias - Taqman probes design'
author:
  - name: Sylvain Moinard
    institute: leca
    email: sylvain.moinard@univ-grenoble-alpes.fr
    correspondence: true
  - name: Didier Piau
    institute: ijf
  - name: Frédéric Laporte
    institute: leca
  - name: Delphine Rioux
    institute: leca
  - name: Pierre Taberlet
    institute: leca
  - name: Christelle Gonindard-Melodelima
    institute: leca
  - name: Eric Coissac
    institute: leca
    email: eric.coissac@metabarcoding.org 
    correspondence: true
institute:
  - leca: Univ. Grenoble-Alpes, Univ. Savoie Mont Blanc, CNRS, LECA, 38000, Grenoble, France.
  - ijf: Univ. Grenoble-Alpes, CNRS, Institut Joseph Fourier, 38000, Grenoble, France.
output:
  html_document
---


## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(purrr)
library(readxl)
library(stringr)
```

## Taqman probe design

This file summarises the steps to design the probes used in the analysis of the paper.

Last edit: 29/05/2024

## probes design

Criteria :

- Length from 15 to 30
- First G as far as possible in the complementary sequence
- Minimal number of G in the complementary sequence
- Difference between Tm_sa(probe) and Tm_sa(primer) (min(F, R)) : about 10°C
- GC content
- Longest homopolymer
- No annealing with non-target sequence

Melting temperature (salt adjusted) : Oligo Calc
http://biotools.nubic.northwestern.edu/OligoCalc.html

```{r functions}
#Position of first G in the probe, starting at position i 
firstG <- function(i, sequence){
  indexG <- str_locate_all(pattern = "c", tolower(sequence))[[1]][,1] #complementary strand
  indexG[which(indexG >= i)[1]]-i+1
}

len_min <- 15
len_max <- 30

Tm_sa <- function(sequence){
  # TM salt adjusted of the probe
  # Treat R, Y, etc as AT
  Na <- 50e-3
  seql <- tolower(sequence)
  100.5 + (41 * (str_count(seql, "g")+str_count(seql, "c"))/nchar(seql)) - (820/nchar(seql)) + 16.6*log10(Na)
}

eval_taqman <- function(sequence, primer = "GGGCAATCCTGAGCCAA"){
  # Evaluate a candidate probe
  # Default primer: Forward Sper01
  res <- data.frame(Sequence = sequence,
             lenght = nchar(sequence),
             Tm_sa = Tm_sa(sequence),
             diffTmsa = Tm_sa(sequence) - Tm_sa(primer), #Sper01
             first_G = firstG(1, sequence),
             number_G = str_count(tolower(sequence), "c"), #complementary
             CG_content = (str_count(tolower(sequence), "g")+
                             str_count(
                               tolower(sequence), "c"))/nchar(sequence),
             sequence_probe = chartr("atgc","tacg",sequence),
             longest_homop = nchar(str_extract_all(
               sequence, "(.)\\1+")%>%
                 map_chr(~.x[which.max(nchar(.x))]))
    )
  rownames(res) <- NULL
  res
}
```

## Open barcode data Sper01

```{r data}
barcodes <- read_excel("data/plants_info/new_plant_positive_control.xlsx")[,c(2,4:6)]
names(barcodes) <- c("species_name", "sequence", "length", "gc_content")

barcodes[barcodes$species_name == "Salvia_pratensis",]$sequence <- "atcctgttttctcaaaacaaaggttcaaaaaacgaaaaaaaaaaag"

barcodes <- barcodes %>%
  filter(!species_name %in%
           c("Taxus_baccata", "Rumex_acetosa", "Picea_abies"))
#Species not used in this experiment
```

## Sper01 primers info

Sper01 : Tm_sa = 54.8°C (F) et 62.1°C (R)

So Tm_sa of probe: around 64°C

```{r sper01}
#Sper01 F : GGGCAATCCTGAGCCAA
Tm_sa("GGGCAATCCTGAGCCAA")

#Sper01 R : GAGCCAAGAGTCTCTGCACCTATC
Tm_sa("CCATTGAGTCTCTGCACCTATC")
```

## Design

Carpinus betulus: long barcode ; two probes

Capsella_bursa-pastoris : higher GC content

Fraxinus excelsior: short barcode


```{r test probes}
taqman_export <- NULL
barcodes$compl <- chartr("atgc","tacg",barcodes$sequence)
```

### Carpinus_betulus

```{r test probes Cbe}
i_Cbe <- which(barcodes$species_name == "Carpinus_betulus")
# nchar(barcodes[i_Cbe,]$sequence)
# barcodes[i_Cbe,]$compl

# str_locate_all(pattern = "g", barcodes[i_Cbe,]$sequence)[[1]][,1]

#Position of first G according to starting base
G1 <- sapply(1:nchar(barcodes[i_Cbe,]$compl),
             function(x) firstG(x, barcodes[i_Cbe,]$compl))
# which.max(G1)

complemCbeA <- substr(barcodes[i_Cbe,]$compl, 1, 30)

taqman_CbeA <- eval_taqman(complemCbeA)
taqman_CbeA$Start_position <- 1
taqman_CbeA$End_position <- 30
taqman_CbeA

complemCbeB <- substr(barcodes[i_Cbe,]$compl, 32, 61)

taqman_CbeB <- eval_taqman(complemCbeB)
taqman_CbeB$Start_position <- 32
taqman_CbeB$End_position <- 61
taqman_CbeB
```

### Capsella_bursa-pastoris

```{r test probes Cbp}
i_Cbp <- which(barcodes$species_name == "Capsella_bursa-pastoris")
# nchar(barcodes[i_Cbp,]$sequence)
# barcodes[i_Cbp,]$compl

# str_locate_all(pattern = "g", barcodes[i_Cbp,]$sequence)[[1]][,1]

#Position of first G according to starting base
G1 <- sapply(1:nchar(barcodes[i_Cbp,]$compl),
             function(x) firstG(x, barcodes[i_Cbp,]$compl))
# which.max(G1)

complemCbp <- substr(barcodes[i_Cbp,]$compl, which.max(G1)[1],
                    which.max(G1)[1]+24)

taqman_Cbp <- eval_taqman(complemCbp)
taqman_Cbp$Start_position <- which.max(G1)[1]
taqman_Cbp$End_position <- which.max(G1)[1]+24
taqman_Cbp
```

###Fraxinus_excelsior

```{r test probes Fex}
i_Fex <- which(barcodes$species_name == "Fraxinus_excelsior")
# nchar(barcodes[i_Fex,]$sequence)
# barcodes[i_Fex,]$compl

# str_locate_all(pattern = "g", barcodes[i_Fex,]$sequence)[[1]][,1]

#Position of first G according to starting base
G1 <- sapply(1:nchar(barcodes[i_Fex,]$compl),
             function(x) firstG(x, barcodes[i_Fex,]$compl))
# which.max(G1)

complemFex <- substr(barcodes[i_Fex,]$compl, which.max(G1)[1],
                    which.max(G1)[1]+29)

taqman_Fex <- eval_taqman(complemFex)
taqman_Fex$Start_position <- which.max(G1)[1]
taqman_Fex$End_position <- which.max(G1)[1]+29
taqman_Fex
```

